ExpressionView file format

### 2010-05-05 RExpressionView/inst/doc/ExpressionView.format.Rnw
EV: XML format vignette updates.

\setlength{\parindent}{2em}

@@

ExpressionView is an interactive visualization tool for biclusters in
gene expression data. The software has two parts. The first part is an
ordering algorithm, written in GNU R, thay shuffles the rows and
columns of the gene expression matrix to make (potentially
overlapping) biclusters more visible. The second part is the
interactive tool, written in Adobe Flex. This runs in an Adobe Flash
enabled web browser. The user can export the ordered gene expression
matrix, with additional meta-data to a data file, that can be openened
by the Adobe Flash application. In this document we briefly discuss
the format of this data file.

@@

The EVF data file, used by ExpressionView, is a standard XML file. The
R package contains an XML Schema file that describes the exact
format. In the following we will show this schema file and explain its
parts step by step, while also showing samples from an example EVF
data file. Parts of the schema file appear in blue boxes, EVF file
code snipplets are in green boxes.

@@

  <xsd:annotation>
    <xsd:documentation xml:lang="en">
     ExpressionView file format schema, version 1.1.
     ExpressionView is a tool to visualize modules (biclusters) in
     gene expression data. Please see http://www.unil.ch/cbg for
     details. Copyright 2010 UNIL DGM Computational Biology Group
    </xsd:documentation>
  </xsd:annotation>
\end{schema}
This is the 1.1 version of the EVF file. ExpressionView can also read
the older 1.0 versions.

\begin{schema}
<xsd:element name="evf" type="evfType" />
\end{schema}
An EVF file contains a single \texttt{<evf>} tag, which has a following
parts:
\begin{schema}
<xsd:complexType name="evfType">
  <xsd:sequence>
    <xsd:element name="summary" type="SummaryType" />
    <xsd:element name="experimentdata" type="ExperimentDataType" />
    <xsd:element name="genes" type="GenesType" />
    <xsd:element name="samples" type="SamplesType" />
    <xsd:element name="modules" type="ModulesType" />
    <xsd:element name="data" type="xsd:base64Binary" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}

@@

\begin{schema}  
<xsd:complexType name="SummaryType">
  <xsd:sequence>
    <xsd:element name="description" minOccurs="0" type="xsd:string" />
    <xsd:element name="version" type="xsd:string" />
    <xsd:element name="dataorigin" minOccurs="0" type="xsd:string" />
    <xsd:element name="xaxislabel" minOccurs="0" type="xsd:string" />
    <xsd:element name="yaxislabel" minOccurs="0" type="xsd:string" />
    <xsd:element name="nmodules" type="xsd:nonNegativeInteger" />
    <xsd:element name="ngenes" type="xsd:positiveInteger" />
    <xsd:element name="nsamples" type="xsd:positiveInteger" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<summary>} tag optionally contains the description of the
data file (\texttt{<description>} tag), this is displayed above the main
gene expression window. The \texttt{<version>} tag is required must be
\texttt{1.1} for files of the format we are discussing here. The
\texttt{<dataorigin>} tag is optional and has the value \texttt{eisa} for
modules generated with the ISA algorithm and exported using the
ExpressionView R package. \texttt{<xaxislabel>} and \texttt{<yaxislabel>}
are optional axis labels for the gene expression plot. The last three
tags (\texttt{<nmodules>}, \texttt{<ngenes>} and \texttt{<nsamples>}) are
required and give the number of modules, number of genes and number of
samples in the data set.

@@

The \texttt{<experimentdata>} tag is next, this contains the
experiment meta-data.
\begin{schema}
<xsd:complexType name="ExperimentDataType">
  <xsd:all>
    <xsd:element name="title" minOccurs="0" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="lab" minOccurs="0" type="xsd:string" />
    <xsd:element name="abstract" minOccurs="0" type="xsd:string" />
    <xsd:element name="url" minOccurs="0" type="xsd:anyURI" />
    <xsd:element name="annotation" minOccurs="0" type="xsd:string" />
    <xsd:element name="organism" minOccurs="0" type="xsd:string" />
  </xsd:all>
</xsd:complexType>
\end{schema}
All fields are optional here: \texttt{<title>}, \texttt{<name>},
\texttt{<lab>}, \texttt{<abstract>}, \texttt{<url>},
\texttt{<annotation>}, \texttt{<organism>}. They are collected and
shown in the \emph{Experiment} tab in the interactive ExpressionView
viewer. The \texttt{<annotation>} tag shoudl contain the name of the
chip on which the experiment was performed. If the \texttt{<organism>}
tag contains \emph{Homo sapiens}, then ExpressionView links genes with
the Gene Cards homepage, for other organism is uses Entrez Gene. The
tags within \texttt{experimentdata} can be specified in arbitrary
order.

@@

The next tag within \texttt{<evf>} is \texttt{<genes>}:
\begin{schema}  
<xsd:complexType name="GenesType">
  <xsd:sequence>
    <xsd:element name="genetags" type="GeneTagsType" />
    <xsd:element name="gene" maxOccurs="unbounded" type="GeneType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
It has two main parts, a \texttt{<genetags>} tag first, and then a
number of \texttt{<gene>} tags, one for each gene in the data set.

Before we continue with the gene tags, we define a new tag type, that
is required to include a \texttt{name} attribute. We call this type
\texttt{ExtraType}. This tag type can be used to add any kind of meta
data to the table in the \emph{Genes} tab in ExpressionView.
\begin{schema}  
<xsd:complexType name="ExtraType">
  <xsd:simpleContent>
    <xsd:extension base="xsd:string">
      <xsd:attribute name="name" type="xsd:string" use="required" />
    </xsd:extension>
  </xsd:simpleContent>
</xsd:complexType>
\end{schema}

The \texttt{<genetags>} tag has one required and many optional
subtags:
\begin{schema}  
<xsd:complexType name="GeneTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="symbol" minOccurs="0" type="xsd:string" />
    <xsd:element name="entrezid" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<id>} must be an integer number, that starts with one and
goes up to the number of genes. \texttt{<name>} is typically the name
of the probe-set, but be used for other purposes as well,
\texttt{<symbol>} is typically the canonical gene
symbol. \texttt{<entrezid>} is the Entrez gene id. If a given
probeset does not map to a known gene, then \texttt{<symbol>} and
\texttt{<entrezid>} can be set to \texttt{NA}. Finally, the gene tags
might contain any number number of \texttt{<x>} tags, each defining an
additional column in the \emph{Genes} tab in ExpressionView. These
tags must have a \texttt{name} attribute, which is refered by the
individual genes. Here is an example gene tags section:
\begin{XML}
<genetags>
  <id>#</id>
  <name>Name</name>
  <symbol>Symbol</symbol>
  <entrezid>EntrezID</entrezid>
  <x name="chr">Chromosome</x>
</genetags>
\end{XML}
Here we define an extra column for the gene table, this will give the
chromosome on which the gene is located.

After the gene tags, we have a typically large number of genes in the file:
\begin{schema}
<xsd:complexType name="GeneType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="symbol" minOccurs="0" type="xsd:string" />
    <xsd:element name="entrezid" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType"/>
  </xsd:sequence>
</xsd:complexType>
\end{schema}
These must have exactly the same subtags as the ones given in the gene
tags section. For example:
\begin{XML}
<gene>
  <id>1</id>
  <name>33500_i_at</name>
  <symbol>NA</symbol>
  <entrezid>NA</entrezid>
  <x name="chr">NA</x>
</gene>
<gene>
  <id>2</id>
  <name>40990_at</name>
  <symbol>TSPAN5</symbol>
  <entrezid>10098</entrezid>
  <x name="chr">4</x>
</gene>
...
\end{XML}
Observe, how the extra tag for the chromosomes is referenced here.

@@

The samples and their associated data have a similar format to the
genes:
\begin{schema}  
<xsd:complexType name="SamplesType">
  <xsd:sequence>
    <xsd:element name="sampletags" type="SampleTagsType" />
    <xsd:element name="sample" maxOccurs="unbounded" 
                 type="SampleType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<samples>} tag has two parts, the first defines the sample
meta data entries and the second part contains one \texttt{<sample>}
tag for each sample in the data set.

\begin{schema}  
<xsd:complexType name="SampleTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<id>} and \texttt{<name>} sample tags are
required. \texttt{<id>} contains numeric ids, from one, up to the
number of samples in the data set. \texttt{<name>} is an arbitrary
string, it is typically the sample name in the R ExpressionSet object,
that was used to find the biclusters. Similarly to the genes, any
number of additional \texttt{<x>} tags can be added, each with a
unique \texttt{name} attribute, to define additional information about
the samples.

Here is an example \texttt{<sampletags>} tag:
\begin{XML}
<sampletags>
  <id>#</id>
  <name>Name</name>
  <x name="diagnosis"> Date of diagnosis</x>
  <x name="sex"> Gender of the patient</x>
  <x name="age"> Age of the patient at entry</x>
  <x name="BT"> does the patient have B-cell or T-cell ALL</x>
</sampletags>
\end{XML}

@@

\begin{schema}  
<xsd:complexType name="ModuleTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="iterations" minOccurs="0" type="xsd:string" />
    <xsd:element name="oscillation" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_row" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_col" minOccurs="0" type="xsd:string" />
    <xsd:element name="freq" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob_limit" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
		 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
There is only one tag required in \texttt{<moduletags>}, the
\texttt{<id>}, which is a numeric id, going from 1 to the number of
modules. The rest is optional, and it is also possible to create any
kind of extra tags with the \texttt{<x>} tag. \texttt{<name>} is an
arbitrary character string. The other optional tags are
mainly defined to modules coming from the ISA algorithm:
\texttt{<iterations>} gives the number of ISA iterations needed to
find the module, \texttt{<oscillation>} specifies the oscillation
cycle length for oscillating modules. It is mostly zero, meaning that
the module does not oscillate. \texttt{<thr\_row>} is the ISA gene
threshold, \texttt{<thr\_col>} is the ISA condition
threshold. \texttt{<freq>} is the number of ISA seeds that converged
to the module, \texttt{<rob>} is its robustness
score. \texttt{<rob\_limit>} is the robustness threshold that was used
to filter the module.

Here is an example \texttt{<moduletags>} tag:
\begin{XML}
<moduletags>
  <id>#</id>
  <name>Name</name>
  <iterations>iterations</iterations>
  <oscillation>oscillation</oscillation>
  <thr_row>thr.row</thr_row>
  <thr_col>thr.col</thr_col>
  <freq>freq</freq>
  <rob>rob</rob>
  <rob_limit>rob.limit</rob_limit>
</moduletags>
\end{XML}

The optional \texttt{<gotags>} and \texttt{<keggtags>} tags are
discussed in the next subsection, let us now see how the data for a
single module looks like:
\begin{schema}
<xsd:complexType name="ModuleType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="iterations" minOccurs="0" type="xsd:string" />
    <xsd:element name="oscillation" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_row" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_col" minOccurs="0" type="xsd:string" />
    <xsd:element name="freq" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob_limit" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
		 type="ExtraType" />
    <xsd:element name="containedgenes" type="integerList" />
    <xsd:element name="genescores" type="doubleList" />
    <xsd:element name="containedsamples" type="integerList" />
    <xsd:element name="samplescores" type="doubleList" />
    <xsd:element name="intersectingmodules" type="integerList" />
    <xsd:element name="gos" minOccurs="0" type="GosType" />
    <xsd:element name="keggs" minOccurs="0" type="KeggsType" />
  </xsd:sequence>
</xsd:complexType>  
\end{schema}
The first couple of fields refer to the ones defined in the
\texttt{<moduletags>} section above. The others are:
\texttt{<containedgenes>}, a space separated list of gene ids, for the
genes that are included in the module; \texttt{<genescores>}, the gene
scores of these genes, in the order of the list in
\texttt{<containedgenes>}; \texttt{<containedsamples>}, the ids of the
samples in the module; \texttt{<samplescores>}, the scores for these
samples; \texttt{<intersectingmodules>}, the ids of the modules that
have an overlap with the given module. \texttt{<gos>} and
\texttt{<keggs>} are optional and describe the Gene Ontology and KEGG
pathway enrichment of the given module, see their details in the next
subsection.

@@

\subsection{Gene Ontology and KEGG pathway enrichment}

EVF files can optionally contain Gene Ontology (GO) and/or KEGG
pathway enrichment calculation results. In this case
\texttt{<moduletags>} is followed by a \texttt{<gotags>} and/or a
\texttt{keggtags} tag:
\begin{schema}
<xsd:complexType name="GoTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="go" type="xsd:string" />
    <xsd:element name="term" type="xsd:string" />
    <xsd:element name="ontology" type="xsd:string" />
    <xsd:element name="pvalue" type="xsd:string" />
    <xsd:element name="oddsratio" type="xsd:string" />
    <xsd:element name="expcount" type="xsd:string" />
    <xsd:element name="count" type="xsd:string" />
    <xsd:element name="size" type="xsd:string" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
Here \texttt{<id>} is a numeric id, starting from one, up to the total
number of enriched GO categories; \texttt{<go>} is the GO id;
\texttt{term} is the GO term; \texttt{ontology} is the GO ontology of
the term, possible values: \texttt{BP}, \texttt{CC}, \texttt{MF};
\texttt{pvalue} is the enrichment $p$-value; \texttt{oddsratio} is the
odds ratio; \texttt{expcount} is the expected number of genes in the
intersection, by chance; \texttt{count} is the number of genes in the
intersection; \texttt{size} is the size of the GO term, i.e. the
number of genes (in the current gene universe) that are annotated with
the term.

@@

Here is part of an EVF file with the \texttt{<gotags>} and
\texttt{<keggtags>} tags:
\begin{XML}
<gotags>
  <id>#</id>
  <go>GO</go>
  <term>Term</term>
  <ontology>Ontology</ontology>
  <pvalue>PValue</pvalue>
  <oddsratio>OddsRatio</oddsratio>
  <expcount>ExpCount</expcount>
  <count>Count</count>
  <size>Size</size>
</gotags>
<keggtags>
  <id>#</id>
  <kegg>KEGG</kegg>
  <pathname>Path Name</pathname>
  <pvalue>PValue</pvalue>
  <oddsratio>OddsRatio</oddsratio>
  <expcount>ExpCount</expcount>
  <count>Count</count>
  <size>Size</size>
</keggtags>
\end{XML}

@@

The subtags within a \texttt{<go>} or a \texttt{<kegg>} tag refer to
the tags already listed in the \texttt{<gotags>} and
\texttt{<keggtags>} tags:
\begin{schema}
<xsd:complexType name="go">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:positiveInteger" />
    <xsd:element name="go" type="xsd:string" />
    <xsd:element name="ontology" type="xsd:string" />
    <xsd:element name="pvalue" type="xsd:double" />
    <xsd:element name="oddsratio" type="xsd:double" />
    <xsd:element name="expcount" type="xsd:double" />
    <xsd:element name="count" type="xsd:nonNegativeInteger" />
    <xsd:element name="size" type="xsd:nonNegativeInteger" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}

@@

Please see the ExpressionView homepage at
\url{http://www.unil.ch/cbg/ExpressionView} for more infomation, the
schema file and example data file.

>>>>

\setlength{\parindent}{12pt}

@@

ExpressionView is an interactive visualization tool for biclusters in
gene expression data. The software has two parts. The first part is an
ordering algorithm, written in GNU R, that reorders the rows and
columns of the gene expression matrix to make (potentially
overlapping) biclusters more visible. The second part is the
interactive tool, written in Adobe Flex. It runs in an Adobe Flash
enabled web browser. The user can export the ordered gene expression
matrix, with additional meta-data from R to a data file, that can be
openened by the Adobe Flash application. In this document we briefly
discuss the format of this data file.

@@

The EVF data file, used by ExpressionView, is a standard XML file. The
R package contains an XML Schema file that describes the exact
format. In the following we will show this schema file and explain its
parts step by step, while also showing samples from an example EVF
data file. Parts of the schema file appear in blue boxes, EVF file
code snipplets are in boxes with green frames.

@@

  <xsd:annotation>
    <xsd:documentation xml:lang="en">
     ExpressionView file format schema, version 1.1.
     ExpressionView is a tool to visualize modules (biclusters) in
     gene expression data. Please see http://www.unil.ch/cbg for
     details. Copyright 2010 UNIL DGM Computational Biology Group
    </xsd:documentation>
  </xsd:annotation>
\end{schema}
This is the 1.1 version of the EVF file. ExpressionView can also read
the older 1.0 version.

\begin{schema}
<xsd:element name="evf" type="evfType" />
\end{schema}
An EVF file contains a single \texttt{<evf>} tag. It has the following
parts:
\begin{schema}
<xsd:complexType name="evfType">
  <xsd:sequence>
    <xsd:element name="summary" type="SummaryType" />
    <xsd:element name="experimentdata" type="ExperimentDataType" />
    <xsd:element name="genes" type="GenesType" />
    <xsd:element name="samples" type="SamplesType" />
    <xsd:element name="modules" type="ModulesType" />
    <xsd:element name="data" type="xsd:base64Binary" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}

@@

This is the type of the \texttt{<summary>} tag:
\begin{schema}  
<xsd:complexType name="SummaryType">
  <xsd:sequence>
    <xsd:element name="description" minOccurs="0" type="xsd:string" />
    <xsd:element name="version" type="xsd:string" />
    <xsd:element name="dataorigin" minOccurs="0" type="xsd:string" />
    <xsd:element name="xaxislabel" minOccurs="0" type="xsd:string" />
    <xsd:element name="yaxislabel" minOccurs="0" type="xsd:string" />
    <xsd:element name="nmodules" type="xsd:nonNegativeInteger" />
    <xsd:element name="ngenes" type="xsd:positiveInteger" />
    <xsd:element name="nsamples" type="xsd:positiveInteger" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<summary>} tag optionally contains the description of the
data file (\texttt{<description>} tag), this is displayed above the main
gene expression window in ExpressionView. The \texttt{<version>} tag
must be \texttt{1.1} for files for the format we are discussing here. The
\texttt{<dataorigin>} tag is optional and has the value `\texttt{eisa}' for
modules generated with the ISA algorithm~\cite{bergmann03} and
exported using the ExpressionView R package. \texttt{<xaxislabel>} and
\texttt{<yaxislabel>} are optional axis labels for the gene expression
plot. The last three tags (\texttt{<nmodules>}, \texttt{<ngenes>} and
\texttt{<nsamples>}) are required and give the number of modules,
number of genes and number of samples in the data set.

@@

The \texttt{<experimentdata>} tag is next, this contains the
experiment meta-data. It has the following parts:
\begin{schema}
<xsd:complexType name="ExperimentDataType">
  <xsd:all>
    <xsd:element name="title" minOccurs="0" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="lab" minOccurs="0" type="xsd:string" />
    <xsd:element name="abstract" minOccurs="0" type="xsd:string" />
    <xsd:element name="url" minOccurs="0" type="xsd:anyURI" />
    <xsd:element name="annotation" minOccurs="0" type="xsd:string" />
    <xsd:element name="organism" minOccurs="0" type="xsd:string" />
  </xsd:all>
</xsd:complexType>
\end{schema}
All fields are optional here: \texttt{<title>}, \texttt{<name>},
\texttt{<lab>}, \texttt{<abstract>}, \texttt{<url>},
\texttt{<annotation>}, \texttt{<organism>}. They are collected and
shown in the \emph{Experiment} tab in the interactive ExpressionView
viewer. The \texttt{<annotation>} tag should contain the name of the
chip on which the experiment was performed. If the \texttt{<organism>}
tag contains `\texttt{Homo sapiens}', then ExpressionView links genes with
the Gene Cards homepage, for other organism is uses Entrez Gene. The
tags within \texttt{experimentdata} can be specified in arbitrary
order.

@@

The next tag within \texttt{<evf>} is \texttt{<genes>}, its type os
defined as
\begin{schema}  
<xsd:complexType name="GenesType">
  <xsd:sequence>
    <xsd:element name="genetags" type="GeneTagsType" />
    <xsd:element name="gene" maxOccurs="unbounded" type="GeneType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
It has two main parts, a \texttt{<genetags>} tag first, and then a
number of \texttt{<gene>} tags, one for each gene in the data set.

Before we continue with the gene tags, we define a new tag type, that
is required to contain a \texttt{name} attribute. We call this type
\texttt{ExtraType}. This tag type can be used to add any kind of meta
data to the table in the \emph{Genes} tab in ExpressionView.
\begin{schema}  
<xsd:complexType name="ExtraType">
  <xsd:simpleContent>
    <xsd:extension base="xsd:string">
      <xsd:attribute name="name" type="xsd:string" use="required" />
    </xsd:extension>
  </xsd:simpleContent>
</xsd:complexType>
\end{schema}

The \texttt{<genetags>} tag has one required and many optional
subtags:
\begin{schema}  
<xsd:complexType name="GeneTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="symbol" minOccurs="0" type="xsd:string" />
    <xsd:element name="entrezid" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<id>} must be an integer number, that starts with one and
goes up to the number of genes. It is used as an id that is refered
from other tags in the document. \texttt{<name>} is typically the name
of the probe-set, but it can be used for other purposes as well,
\texttt{<symbol>} is typically the canonical gene
symbol. \texttt{<entrezid>} is the Entrez gene id. If a given
probeset does not map to a known gene, then \texttt{<symbol>} and
\texttt{<entrezid>} can be set to \texttt{NA}. Finally, the gene tags
might contain any number of \texttt{<x>} tags, each defining an
additional column in the \emph{Genes} tab in ExpressionView. These
tags must have a \texttt{name} attribute, which is refered by the
tags of the individual genes. Here is an example gene tags section:
\begin{XML}
<genetags>
  <id>#</id>
  <name>Name</name>
  <symbol>Symbol</symbol>
  <entrezid>EntrezID</entrezid>
  <x name="chr">Chromosome</x>
</genetags>
\end{XML}
We define an extra column for the gene table, this will give the
chromosome on which the gene is located.

After the gene tags, we have a typically large number of genes in the
file. The definition of the \texttt{<gene>} tag:
\begin{schema}
<xsd:complexType name="GeneType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="symbol" minOccurs="0" type="xsd:string" />
    <xsd:element name="entrezid" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType"/>
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<gene>} tags must have exactly the same subtags as the
ones given in the gene tags section. For example, assuming the
\texttt{<genetags>} example above we can have:
\begin{XML}
<gene>
  <id>1</id>
  <name>33500_i_at</name>
  <symbol>NA</symbol>
  <entrezid>NA</entrezid>
  <x name="chr">NA</x>
</gene>
<gene>
  <id>2</id>
  <name>40990_at</name>
  <symbol>TSPAN5</symbol>
  <entrezid>10098</entrezid>
  <x name="chr">4</x>
</gene>
...
\end{XML}
Observe, how the extra tag for the chromosomes is referenced here.

@@

The samples and their associated data have a format similar to the
genes:
\begin{schema}  
<xsd:complexType name="SamplesType">
  <xsd:sequence>
    <xsd:element name="sampletags" type="SampleTagsType" />
    <xsd:element name="sample" maxOccurs="unbounded" 
                 type="SampleType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<samples>} tag has two parts, the first defines the sample
meta data entries and the second part contains one \texttt{<sample>}
tag for each sample in the data set.

\begin{schema}  
<xsd:complexType name="SampleTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<id>} and \texttt{<name>} sample tags are
required. \texttt{<id>} contains numeric ids, from one, up to the
number of samples in the data set. These are referenced by other tags,
e.g. the ones that define the modules. \texttt{<name>} is an arbitrary
string, it is typically the sample name in the R ExpressionSet object,
that was used to find the biclusters. Similarly to the genes, any
number of additional \texttt{<x>} tags can be added, each with a
unique \texttt{name} attribute, to define additional information about
the samples.

Here is an example \texttt{<sampletags>} tag:
\begin{XML}
<sampletags>
  <id>#</id>
  <name>Name</name>
  <x name="diagnosis"> Date of diagnosis</x>
  <x name="sex"> Gender of the patient</x>
  <x name="age"> Age of the patient at entry</x>
  <x name="BT"> does the patient have B-cell or T-cell ALL</x>
</sampletags>
\end{XML}
We defined four extra tags for the samples in this example.

@@

The type of \texttt{<moduletags>}:
\begin{schema}  
<xsd:complexType name="ModuleTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="iterations" minOccurs="0" type="xsd:string" />
    <xsd:element name="oscillation" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_row" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_col" minOccurs="0" type="xsd:string" />
    <xsd:element name="freq" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob_limit" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
		 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
There is only one tag required in \texttt{<moduletags>}, the
\texttt{<id>}, which is a numeric id, going from 1 to the number of
modules. The rest is optional, and it is also possible to create any
kind of extra tags with the \texttt{<x>} tag. \texttt{<name>} is an
arbitrary character string. The other optional tags are
mainly defined to modules coming from the ISA algorithm:
\texttt{<iterations>} gives the number of ISA iterations needed to
find the module, \texttt{<oscillation>} specifies the oscillation
cycle length for oscillating modules. It is mostly zero, meaning that
the module does not oscillate. \texttt{<thr\_row>} is the ISA gene
threshold, \texttt{<thr\_col>} is the ISA condition
threshold. \texttt{<freq>} is the number of ISA seeds that converged
to the module, \texttt{<rob>} is its robustness
score. \texttt{<rob\_limit>} is the robustness threshold that was used
to filter the module.

Here is an example \texttt{<moduletags>} tag:
\begin{XML}
<moduletags>
  <id>#</id>
  <name>Name</name>
  <iterations>Iterations</iterations>
  <oscillation>Oscillation cycle</oscillation>
  <thr_row>Gene threshold</thr_row>
  <thr_col>Column threshold</thr_col>
  <freq>Frequency</freq>
  <rob>Robustness</rob>
  <rob_limit>Robustness limit</rob_limit>
</moduletags>
\end{XML}

The optional \texttt{<gotags>} and \texttt{<keggtags>} tags are
discussed in Section~\ref{sec:gokegg}, let us now see how the data for a
single module looks like:
\begin{schema}
<xsd:complexType name="ModuleType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="iterations" minOccurs="0" type="xsd:string" />
    <xsd:element name="oscillation" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_row" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_col" minOccurs="0" type="xsd:string" />
    <xsd:element name="freq" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob_limit" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
		 type="ExtraType" />
    <xsd:element name="containedgenes" type="integerList" />
    <xsd:element name="genescores" type="doubleList" />
    <xsd:element name="containedsamples" type="integerList" />
    <xsd:element name="samplescores" type="doubleList" />
    <xsd:element name="intersectingmodules" type="integerList" />
    <xsd:element name="gos" minOccurs="0" type="GosType" />
    <xsd:element name="keggs" minOccurs="0" type="KeggsType" />
  </xsd:sequence>
</xsd:complexType>  
\end{schema}
The first couple of fields refer to the ones defined in the
\texttt{<moduletags>} section above. The others are:
\texttt{<containedgenes>}, a space separated list of gene ids, for the
genes that are included in the module; \texttt{<genescores>}, the gene
scores of these genes, in the order of the list in
\texttt{<containedgenes>}; \texttt{<containedsamples>}, the ids of the
samples in the module; \texttt{<samplescores>}, the scores for these
samples; \texttt{<intersectingmodules>}, the ids of the modules that
have an overlap with the given module. \texttt{<gos>} and
\texttt{<keggs>} are optional and describe the Gene Ontology and KEGG
pathway enrichment of the given module, see their details in Section~\ref{sec:gokegg}.

@@

\subsection{Gene Ontology and KEGG pathway enrichment}%
\label{sec:gokegg}

EVF files can optionally contain Gene Ontology (GO) and/or KEGG
pathway enrichment calculation results. In this case
\texttt{<moduletags>} is followed by a \texttt{<gotags>} and/or a
\texttt{<keggtags>} tag. The definition of \texttt{<gotags>}:
\begin{schema}
<xsd:complexType name="GoTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="go" type="xsd:string" />
    <xsd:element name="term" type="xsd:string" />
    <xsd:element name="ontology" type="xsd:string" />
    <xsd:element name="pvalue" type="xsd:string" />
    <xsd:element name="oddsratio" type="xsd:string" />
    <xsd:element name="expcount" type="xsd:string" />
    <xsd:element name="count" type="xsd:string" />
    <xsd:element name="size" type="xsd:string" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
\texttt{<id>} is a numeric id, starting from one, up to the total
number of enriched GO categories; \texttt{<go>} is the GO id;
\texttt{<term>} is the GO term; \texttt{<ontology>} is the GO ontology of
the term, its possible values are: \texttt{BP}, \texttt{CC},
\texttt{MF}, standing for \emph{Biological process}, \emph{Cellular
  component} and \emph{Molecular function};
\texttt{<pvalue>} is the enrichment $p$-value; \texttt{<oddsratio>} is the
odds ratio; \texttt{<expcount>} is the expected number of genes in the
intersection, by chance; \texttt{<count>} is the number of genes in the
intersection; \texttt{<size>} is the size of the GO term, i.e. the
number of genes (in the current gene universe) that are annotated with
the enriched term.

@@

Part of an EVF file with the \texttt{<gotags>} and
\texttt{<keggtags>} tags:
\begin{XML}
<gotags>
  <id>#</id>
  <go>GO</go>
  <term>Term</term>
  <ontology>Ontology</ontology>
  <pvalue>PValue</pvalue>
  <oddsratio>OddsRatio</oddsratio>
  <expcount>ExpCount</expcount>
  <count>Count</count>
  <size>Size</size>
</gotags>
<keggtags>
  <id>#</id>
  <kegg>KEGG</kegg>
  <pathname>Path Name</pathname>
  <pvalue>PValue</pvalue>
  <oddsratio>OddsRatio</oddsratio>
  <expcount>ExpCount</expcount>
  <count>Count</count>
  <size>Size</size>
</keggtags>
\end{XML}

@@

The subtags within a \texttt{<go>} or a \texttt{<kegg>} tag refer to
the tags already listed above in the \texttt{<gotags>} and
\texttt{<keggtags>} tags:
\begin{schema}
<xsd:complexType name="go">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:positiveInteger" />
    <xsd:element name="go" type="xsd:string" />
    <xsd:element name="ontology" type="xsd:string" />
    <xsd:element name="pvalue" type="xsd:double" />
    <xsd:element name="oddsratio" type="xsd:double" />
    <xsd:element name="expcount" type="xsd:double" />
    <xsd:element name="count" type="xsd:nonNegativeInteger" />
    <xsd:element name="size" type="xsd:nonNegativeInteger" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}

@@

Please see the ExpressionView homepage at
\url{http://www.unil.ch/cbg/ExpressionView} for more infomation, the
schema file and example data files.
